---
a---
title: "Hypothesis"
author: "Allison Louie"
date: "2024-05-22"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    code_folding: hide
---

```{r setup, include=FALSE}
library(knitr)
library(summarytools) # for summarizing variables
library(tidyverse)
install.packages("BiocManager")
BiocManager::install("Biobase")
library(Biobase)
options(knitr.table.format = "html")
knitr::opts_chunk$set(echo = TRUE)

# folder for analysis
work.dir <- here::here()

# suppress warning messages for final rendering
old.warn <- getOption("warn")
options(warn=-1)
```

# Hypothesis

Prenatal exposure to lifestyle and chemicals are associated with variations in BMI z-score in children at age 6-11 years. This association will be controlled for dietary intake, physical activity, and socioeconomic status during and after pregnancy. The serum metabolomic data would be helpful to look at in terms of long-term metabolic status and systemic health related to chronic exposure effects, which could be linked to weight regulation and obesity.

# Background

Prenatal exposure to phthalates, which are widely recognized as endocrine-disrupting chemicals, has been increasingly associated with adverse health outcomes in children. Research suggests that phthalates may influence body weight regulation and metabolic processes through their hormonal actions, which could predispose children to obesity. A recent study within the Ma'anshan birth cohort highlighted the impact of prenatal exposure to phthalates such as mono-ethyl phthalate (MEP) and di(2-ethylhexyl) phthalate (DEHP) on the trajectory of body mass index (BMI) in girls during the first six years of life. This study found that higher prenatal phthalate concentrations were associated with higher BMI trajectories, suggesting that exposure during pregnancy could play a significant role in shaping future obesity risk (Gao et al., 2022).

Research from the INMA cohort study (Montazeri et al., 2023) has highlighted that exposure to multiple EDCs during pregnancy is associated with varied growth trajectories in children, from birth through childhood. Specifically, persistent chemicals like hexachlorobenzene (HCB), DDE (a breakdown product of DDT), and PCBs have been found to contribute significantly to accelerated BMI gain starting from lower or higher birth sizes. This study suggests that prenatal exposure to these chemicals may predispose children to faster postnatal growth rates, which are linked to higher risks of obesity and other metabolic disorders later in life.

Moreover, a study by Yang et al. (2018) demonstrated that prenatal exposure to phthalates is linked to sex-specific differences in BMI trajectories from birth to adolescence, with third-trimester exposures being particularly critical. These findings underscore the potential for phthalates to disrupt metabolic health during critical windows of prenatal development, emphasizing the need for further exploration into how these exposures affect long-term health outcomes in children.

This study aims to provide further insights into the mechanisms by which prenatal environmental exposures contribute to obesity, potentially guiding public health strategies aimed at reducing these exposures during pregnancy.

# Variables:

### Exposure Variables:

#### Lifestyle:

##### Pregnancy

-   prenatal alcohol

-   cereal consumption during pregnancy (times/week)

-   dairy consumption during pregnancy (times/week)

-   fast food consumption during pregnancy (times/week)

-   fish consumption during pregnancy (times/week)

-   folic acid supplementation during pregnancy

-   fruit consumption during pregnancy (times/week)

-   legume consumption during pregnancy (times/week)

-   meat consumption during pregnancy (times/week)

-   Walking and/or cycling activity during pregnancy (frequency)

-   Exercise or sport activity during pregnancy (frequency)

-   vegetables consumption during pregnancy (times/week)

##### Postnatal

-   Food group: bakery products (hs_cookies + hs_pastries)

-   Food group: beverages (hs_dietsoda+hs_soda)

-   Food group: breakfast cereal (hs_sugarcer+hs_othcer)

-   Drinks a caffeinated or æenergy drink (eg coca-cola, diet-coke, redbull)

-   Food group: dairy (hs_cheese + hs_milk + hs_yogurt+ hs_probiotic+ hs_desert)

-   Visits a fast food restaurant/take away

-   Sum of KIDMED indices, without index9

-   Clean & Over-reporting of Moderate-to-Vigorous Physical Activity (min/day)

-   Eats organic food

-   Food group: processed meat (hs_coldmeat+hs_ham)

-   Eats a æready-made supermarket meal

-   sedentary behaviour (min/day)

-   Food group: bread (hs_darkbread+hs_whbread)

-   Food group: cereal (hs_darkbread + hs_whbread + hs_rice_pasta + hs_sugarcer + hs_othcer + hs_rusks)

-   Food group: fish and seafood (hs_canfish+hs_oilyfish+hs_whfish+hs_seafood)

-   Food group: fruits (hs_canfruit+hs_dryfruit+hs_freshjuice+hs_fruits)

-   Food group: Added fat

-   Food group: meat (hs_coldmeat+hs_ham+hs_poultry+hs_redmeat)

-   Food group: potatoes (hs_frenchfries+hs_potatoes)

-   Food group: sweets (hs_choco + hs_sweets + hs_sugar)

-   Food group: vegetables (hs_cookveg+hs_rawveg)

-   Food group: yogurt (hs_yogurt+hs_probiotic)

-   Total hours of sleep (mean weekdays and night)

#### Chemicals:

##### Pregnancy

-   Arsenic (As) in mother

-   Cadmium (Cd) in mother

-   Cobalt (Co) in mother

-   Caesium (Cs) in mother

-   Copper (Cu) in mother

-   Mercury (Hg) in mother

-   Manganese (Mn) in mother

-   Molybdenum (Mo) in mother

-   Lead (Pb) in mother

-   Dichotomous variable of thallium (Tl) in mother

-   Dichlorodiphenyldichloroethylene (DDE) in mother adjusted for lipids

-   Dichlorodiphenyltrichloroethane (DDT) in mother adjusted for lipids

-   Hexachlorobenzene (HCB) in mother adjusted for lipids

-   Polychlorinated biphenyl-118 (PCB-118) in mother adjusted for lipids

-   Polychlorinated biphenyl-138 (PCB-138) in mother adjusted for lipids

-   Polychlorinated biphenyl-153 (PCB-153) in mother adjusted for lipids

-   Polychlorinated biphenyl-170 (PCB-170) in mother adjusted for lipids

-   Polychlorinated biphenyl-180 (PCB-180) in mother adjusted for lipids

-   Sum of PCBs in mother adjusted for lipids (5 cohorts)

-   Diethyl phosphate (DEP) in mother adjusted for creatinine

-   Diethyl thiophosphate (DETP) in mother adjusted for creatinine

-   Dimethyl phosphate (DMP) in mother adjusted for creatinine

-   Dimethyl thiophosphate (DMTP) in child adjusted for creatinine

-   Polybrominated diphenyl ether-153 (PBDE-153) in mother adjusted for lipids

-   Polybrominated diphenyl ether-47 (PBDE-47) in mother adjusted for lipids

-   Perfluorohexane sulfonate (PFHXS) in mother

-   Perfluorononanoate (PFNA) in mother

-   Perfluorooctanoate (PFOA) in mother

-   Perfluorooctane sulfonate (PFOS) in mother

-   Perfluoroundecanoate (PFUNDA) in mother

-   Bisphenol A (BPA) in mother adjusted for creatinine

-   N-Butyl paraben (BUPA) in mother adjusted for creatinine

-   Ethyl paraben (ETPA) in mother adjusted for creatinine

-   Methyl paraben (MEPA) in mother adjusted for creatinine

-   Oxybenzone (OXBE) in mother adjusted for creatinine

-   Propyl paraben (PRPA) in mother adjusted for creatinine

-   Triclosan (TRCS) in mother adjusted for creatinine

-   Mono benzyl phthalate (MBzP) in mother adjusted for creatinine

-   Mono-2-ethyl 5-carboxypentyl phthalate (MECPP) in mother adjusted for creatinine

-   Mono-2-ethyl-5-hydroxyhexyl phthalate (MEHHP) in mother adjusted for creatinine

-   Mono-2-ethylhexyl phthalate (MEHP) in mother adjusted for creatinine

-   Mono-2-ethyl-5-oxohexyl phthalate (MEOHP) in mother adjusted for creatinine

-   Monoethyl phthalate (MEP) in mother adjusted for creatinine

-   Mono-iso-butyl phthalate (MiBP) in mother adjusted for creatinine

-   Mono-n-butyl phthalate (MnBP) in mother adjusted for creatinine

-   Mono-4-methyl-7-hydroxyoctyl phthalate (OHMiNP) in mother adjusted for creatinine

-   Mono-4-methyl-7-oxooctyl phthalate (OXOMiNP) in mother adjusted for creatinine

-   Sum of DEHP metabolites (µg/g) in mother adjusted for creatinine

-   maternal active Tobacco Smoke pregnancy mean nb cig/day

-   Categorical variable of cotinine in mother

##### Postnatal

-   Arsenic (As) in child

-   Cadmium (Cd) in child

-   Cobalt (Co) in child

-   Caesium (Cs) in child

-   Copper (Cu) in child

-   Mercury (Hg) in child

-   Manganese (Mn) in child

-   Molybdenum (Mo) in child

-   Lead (Pb) in child

-   Dichotomous variable of thallium (Tl) in child

-   Dichlorodiphenyldichloroethylene (DDE) in child adjusted for lipids

-   Dichlorodiphenyltrichloroethane (DDT) in child adjusted for lipids

-   Hexachlorobenzene (HCB) in child adjusted for lipids

-   Polychlorinated biphenyl -118 (PCB-118) in child adjusted for lipids

-   Polychlorinated biphenyl-138 (PCB-138) in child adjusted for lipids

-   Polychlorinated biphenyl-153 (PCB-153) in child adjusted for lipids

-   Polychlorinated biphenyl-170 (PCB-170) in child adjusted for lipids

-   Polychlorinated biphenyl-180 (PCB-180) in child adjusted for lipids

-   Sum of PCBs in child adjusted for lipids (4 cohorts)

-   Diethyl phosphate (DEP) in child adjusted for creatinine

-   Diethyl thiophosphate (DETP) in child adjusted for creatinine

-   Dichotomous variable of dimethyl dithiophosphate (DMDTP) in child

-   Dimethyl phosphate (DMP) in child adjusted for creatinine

-   Polybrominated diphenyl ether-153 (PBDE-153) in child adjusted for lipids

-   Polybrominated diphenyl ether-47 (PBDE-47) in child adjusted for lipids

-   Perfluorohexane sulfonate (PFHXS) in child

-   Perfluorononanoate (PFNA) in child

-   Perfluorooctanoate (PFOA) in child

-   Perfluorooctane sulfonate (PFOS) in child

-   Perfluoroundecanoate (PFUNDA) in child

-   Bisphenol A (BPA) in child adjusted for creatinine

-   N-Butyl paraben (BUPA) in child adjusted for creatinine

-   Ethyl paraben (ETPA) in child adjusted for creatinine

-   Methyl paraben (MEPA) in child adjusted for creatinine

-   Oxybenzone (OXBE) in child adjusted for creatinine

-   Propyl paraben (PRPA) in child adjusted for creatinine

-   Triclosan (TRCS) in child adjusted for creatinine

-   Mono benzyl phthalate (MBzP) in child adjusted for creatinine

-   Mono-2-ethyl 5-carboxypentyl phthalate (MECPP) in child adjusted for creatinine

-   Mono-2-ethyl-5-hydroxyhexyl phthalate (MEHHP) in child adjusted for creatinine

-   Mono-2-ethylhexyl phthalate (MEHP) in child adjusted for creatinine

-   Mono-2-ethyl-5-oxohexyl phthalate (MEOHP) in child adjusted for creatinine

-   Monoethyl phthalate (MEP) in child adjusted for creatinine

-   Mono-iso-butyl phthalate (MiBP) in child adjusted for creatinine

-   Mono-n-butyl phthalate (MnBP) in child adjusted for creatinine

-   Mono-4-methyl-7-hydroxyoctyl phthalate (OHMiNP) in child adjusted for creatinine

-   Mono-4-methyl-7-oxooctyl phthalate (OXOMiNP) in child adjusted for creatinine

-   Sum of DEHP metabolites (µg/g) in child adjusted for creatinine

-   Family affluence score

-   scoial capital: family friends

-   How many people live in your home?

-   social capital: structural

-   Dichotomous variable of cotinine in child

-   Global exposure of the child to ETS (2 categories)

-   Tobacco Smoke status of parents (both)

### Covariates

-   Maternal pre-pregnancy body mass index (kg/m2)

-   Height of the child at 6-11 years old (m)

-   Weight of the child at 6-11 years old (kg)

-   Maternal weight gain during pregnancy (kg)

-   Gestational age at birth (week)

-   Child sex (female / male)

-   Year of birth (2003 to 2009)

-   Maternal age (years)

-   Cohort of inclusion (1 to 6)

-   Maternal education (1: primary school, 2:secondary school, 3:university degree or higher)

-   Are the parents native from the country of the cohort (0: no native parent, 1:only one native parent, 2: both parents native)

-   Parity before index pregnancy (0: nulliparous, 1:primiparous, 2:multiparous)

-   Child age at examination (years)

### Outcome Variable:

Child BMI Z-score measured at age 6-11 years.

# Data Description

Data for this study were derived from the six HELIX cohorts, encompassing over 31,472 mother-child pairs at different pregnancy stages from 1999 to 2010. The analysis focuses on 1,301 mother-children pairs from the HELIX subcohort, which aged 6-11 years for whom complete exposure and outcome data were available. Exposure data included detailed dietary records during pregnancy and concentrations of various chemicals like BPA and PCBs in maternal blood samples. Child BMI Z-scores were calculated based on WHO growth standards. Preliminary descriptive analysis showed a balanced distribution across genders and a wide range of socioeconomic statuses.

```{r load data and coddebook, echo=TRUE}
load(paste0(work.dir, "/HELIX.RData"))
kable(codebook, align="c")
```

## Data Summary for Covariates, Phenotypes, and Exposures {.tabset}

### Data Summary Covariates

```{r covariate summary, results='asis'}
summarytools::view(dfSummary(covariates, style = 'grid', plain.ascii = FALSE, valid.col = FALSE, headings = FALSE), method = "render")
```

### Data Summary Phenotypes

```{r phenotype summary}
summarytools::view(dfSummary(phenotype, style = 'grid', plain.ascii = FALSE, valid.col = FALSE, headings = FALSE), method = "render")
```

### Data Summary Exposures: Outdoor exposures

```{r Outdoor exposures summary, attr.output='style="max-height: 100px;"',}
selectExposures <- codebook$variable_name[codebook$domain=="Outdoor exposures"]
summarytools::view(dfSummary(exposome[,names(exposome) %in% selectExposures], style = 'grid', plain.ascii = FALSE, valid.col = FALSE, headings = FALSE), method = "render")
```

### Data Summary Exposures: Indoor air

```{r Indoor air summary, attr.output='style="max-height: 100px;"',}
selectExposures <- codebook$variable_name[codebook$domain=="Indoor air"]
summarytools::view(dfSummary(exposome[,names(exposome) %in% selectExposures], style = 'grid', plain.ascii = FALSE, valid.col = FALSE, headings = FALSE), method = "render")
```

### Data Summary Exposures: Lifestyles

```{r Lifestyles summary, attr.output='style="max-height: 100px;"',}
selectExposures <- codebook$variable_name[codebook$domain=="Lifestyles"]
summarytools::view(dfSummary(exposome[,names(exposome) %in% selectExposures], style = 'grid', plain.ascii = FALSE, valid.col = FALSE, headings = FALSE), method = "render")
```

### Data Summary Exposures: Chemicals

```{r Chemicals summary, attr.output='style="max-height: 100px;"',}
selectExposures <- codebook$variable_name[codebook$domain=="Chemicals"]
summarytools::view(dfSummary(exposome[,names(exposome) %in% selectExposures], style = 'grid', plain.ascii = FALSE, valid.col = FALSE, headings = FALSE), method = "render")
```

## Proteomic Data

First 10 rows and columns of the proteomic data

```{r proteomic data, echo=TRUE}
load(paste0(work.dir, "/proteome.RData"))
kable(proteome.d[1:10,1:10], align="c", digits=2, format="pipe")


```

## Metabolomic Urine Data

First 10 rows and columns of the metabolomic urine data

```{r metabolomic urine data, echo=TRUE}
load(paste0(work.dir, "/metabol_urine.RData"))
kable(metabol_urine.d[1:10,1:10], align="c", digits=2, format="pipe")


```

## Metabolomic Serum Data

First 10 rows and columns of the metabolomic serum data

```{r metabolomic serum data, echo=TRUE}
load(paste0(work.dir, "/metabol_serum.RData"))
kable(metabol_serum.d[1:10,1:10], align="c", digits=2, format="pipe")
```

## Genomic Data

First 10 rows and columns of the genomic data

```{r genomic data, echo=TRUE}
load(paste0(work.dir, "/genome.RData"))
kable(G[1:10,1:10], align="c", digits=2, format="pipe")
```
